{\sc Matlab}\textsuperscript{\textregistered} is a popular dynamic scientific
programming language. The typical \matlab user is not a software professional;
it is chiefly used among scientists, engineers, and students, and enjoys wide
adoption in large part because of its high level syntax and wide array of
libraries for many problem domains in the sciences. The inexperience of many
\matlab programmers, coupled with the ill-specified and often counterintuitive
semantics of the language, leads to \matlab code in the wild that is difficult
to understand and maintain.

In this thesis, we present McIDE, an integrated development environment for
\matlab programming. McIDE provides tools to help \matlab programmers write
better programs, among them automated refactorings and code navigation features
like "jump to definition". It is also opinionated about \matlab code, and tries
to recognize common anti-patterns and either warn about or eliminate them.

McIDE is built up of several largely independent components wired together by a
thin graphical interface. Some of these components are pre-existing, such a
\matlab parser provided by the \mclab compiler toolkit, and others are
contributions of this thesis, such as a dynamic call graph collection mechanism
for \matlab code, and a layout-preserving code transformation engine.

A theme of McIDE's implementation is reliance on runtime information, since
purely static information is often insufficient if we wish to support the
development of arbitrary \matlab code, including its more dynamic features.
